---
title: "Normalization: A pipeline to analyse time-course gene expression data"
output: rmarkdown::html_vignette
---



```{r echo=FALSE, results="hide"}
# Loading dependencies
library(devtools)
library(limma)
library(splines)
library(stats)
library(moanin)
source("utils.R")

# library(moanin)
```

```{r echo=FALSE}
# Set options
df = 6

# Preprocessing & filtering
is_count = FALSE
take_log = FALSE    
filter_expression = FALSE

# Differential expression analysis
timecourse_contrasts = c("C-K", "C-M")

# Clustering
filter_genes = TRUE
n_genes_to_keep = 5000
n_clusters = 20
percentage_genes_to_label = 0.5
```

```{r echo=FALSE, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  cache=FALSE, autodep=FALSE, warning=FALSE, error=FALSE, message=FALSE,
  echo=FALSE,
  duplicate.label="allow"
)

library(genefilter)
library(dplyr)
library(ggplot2)
library(NMF)
library(reshape2)
library(limma)
```



The paper describes a normalization strategy that I am going to use here:

  - First, normalize probe intensity using background information using the
    "norm-exponential" strategy.
  - Then normalize across micro-arrrays using quantily normalization.

Unfortunately, that means going back to the raw data, which can be downloaded
in the data folder, by clicking on the link specified in the README (it might
take a while…). Untar the main file into a `rawdata` folder.

```{r reading_targets, echo=FALSE, include=FALSE}

targets = readTargets()
RG = read.maimages(
  targets, source="agilent",
  path="data/shoemaker2015/rawdata",
  other.columns="gIsWellAboveBG",
  green.only=TRUE)
```

Before quantile normalization:

```{r echo=FALSE, include=FALSE}
RG = backgroundCorrect(RG, method="normexp")
```

```{r plot_densities, echo=FALSE}
xlim = c(0, 14)
plotDensities(RG, col="black", legend=FALSE)
```

After quantile normalization:

```{r normalize_between_arrays, echo=FALSE}
RG.q = normalizeBetweenArrays(RG, method="quantile")
plotDensities(RG.q, col="black", legend=FALSE)

sample_names = sapply(strsplit(colnames(RG.q), "_"), .subset2,  1)
colnames(RG.q) = sample_names
row.names(RG.q) = make.names(RG$genes$SystematicName, unique=TRUE)
idx_to_keep = which(
  !(RG$genes$Systematic %in% c("DarkCorner", "GE_BrightCorner",
			       "NegativeControl")))

data = RG.q[idx_to_keep,]

# Filter out control probes.
control_probes = data$genes[, "ControlType"] == 1
is_expressed = rowSums(data$other$"gIsWellAboveBG" > 0) >= 4

data = data[!control_probes & is_expressed, ]$E

write.table(data, ".results/quantile_normalized.txt")

meta = read.delim("data/shoemaker2015/mice_meta.tsv", sep="\t")

```

```{r order_genes, echo=FALSE}
# Reorder genes on condition, time, and replicate
ord = order(
  meta$Group,
  meta$Time,
  meta$Replicate)
data = data[, ord]
meta = meta[ord, ]
meta["Time"] = as.factor(unlist(meta["Time"]))
meta["Replicate"] = as.factor(unlist(meta["Replicate"]))
```


## Plot the RLE and PCA

```{r leafRleRaw, fig.width=6}
library(EDASeq)
ylim = c(-2.5,2.5)
plotRLE(exp(data), outline=FALSE,
        whisklty=1, whisklwd=2, col=meta$Replicate, main="RLE, data")
plotRLE(exp(data), outline=FALSE,
        whisklty=1, whisklwd=2, col=meta$Group, main="RLE, data")
plotRLE(exp(data), outline=FALSE,
        whisklty=1, whisklwd=2, col=meta$Time, main="RLE, data")

```
```{r leafPCARaw}
library(ggfortify)
pca_data = prcomp(t(data), rank=5, center=TRUE, scale=TRUE) 
percent_var = round(100 * attr(pca_data, "percentVar"))

autoplot(pca_data, data=meta, colour="Group")
autoplot(pca_data, data=meta, colour="Time")
autoplot(pca_data, data=meta, colour="Replicate")

```

```{r fig.width=6, fig.height=6}
library(NMF)

data_corr = cor(data, method="pearson")
aheatmap(data_corr,
         Colv=NA, Rowv=NA,
         annCol=meta[,
		     c("Group", "Time",
			"Replicate")],
         annLegend=FALSE, 
         main="Replicates\nAfter Normalization")
```

```{r fig.width=6, fig.height=6}
library(NMF)

data_corr = cor(data, method="spearman")
aheatmap(data_corr,
         Colv=NA, Rowv=NA,
         annCol=meta[,
		     c("Group", "Time",
			"Replicate")],
         annLegend=FALSE, 
	 annColors=ann_colors,
         main="Replicates\nAfter Normalization")
```
